### 2025-05-29 chapters/background.tex
updated relevant chapters & sections

\section{Trusted Execution Environments}

@@

@@

>>>>

\section{Trusted Execution Environments}
Trusted Execution Environments (TEEs) represent a class of hardware-assisted secure computing technologies that provide strong guarantees for the confidentiality and integrity of both code and data, even in systems where the underlying operating system or application software may be compromised. TEEs are designed to protect sensitive computations from a wide range of adversaries by establishing isolated execution contexts—commonly referred to as enclaves—through a combination of hardware features and privileged system software.

A TEE isolates a subset of the system’s resources, including CPU registers, memory, and I/O peripherals, and makes them accessible only to trusted applications executing within the enclave boundary. This isolation is enforced by the processor’s privilege mechanisms, memory protection units, or dedicated security monitors, ensuring that unauthorized entities—whether kernel-level malware or user-space processes—cannot access or tamper with enclave-resident assets.

Security within a TEE is achieved by reducing the size and complexity of the Trusted Computing Base (TCB). The TCB encompasses the set of components that must be trusted for the TEE to function securely. A minimal TCB is desirable, as it reduces the likelihood of vulnerabilities and simplifies formal verification and auditing processes. Modern TEEs aim to move as few components as possible into the TCB, while still providing robust isolation and essential services.

Prominent commercial TEE implementations include Intel Software Guard Extensions (SGX) and ARM TrustZone. Intel SGX facilitates the creation of enclaves at the user level, with memory encryption and access control enforced by hardware. ARM TrustZone, in contrast, provides a split-world architecture that partitions the processor into a secure world and a normal world, with secure world execution isolated via monitor calls. While these solutions have been successfully deployed in many real-world systems, their proprietary nature, rigid designs, and opaque implementations inhibit their utility for research, experimentation, and system-level innovation.

In response to these limitations, open-source TEE frameworks such as Keystone have emerged. Keystone is built atop the RISC-V instruction set architecture (ISA), which is itself open, extensible, and amenable to security-centric innovation. RISC-V’s modularity enables the design of TEEs that are both highly customizable and more transparent in their operation. A foundational architectural principle shared by all TEEs, including Keystone, is privilege separation. This concept involves executing the core TCB in a highly privileged execution mode—such as machine mode (M-mode) in RISC-V—while relegating the operating system and general-purpose applications to less privileged levels. This separation is crucial for enforcing isolation policies and preventing untrusted software from influencing the secure execution environment.

By leveraging privilege separation, physical memory protection, and modular TCB design, TEEs provide a powerful framework for secure computation in adversarial environments. Keystone builds upon these principles and extends them with openness and configurability, making it a valuable platform for academic and practical exploration of TEE technologies.

@@

Keystone is an open-source Trusted Execution Environment (TEE) framework designed specifically for the RISC-V architecture. It enables the development of secure enclaves that are customizable, lightweight, and suitable for both research and deployment in resource-constrained or security-critical environments. Keystone’s design reflects a modular approach, emphasizing minimalism in its Trusted Computing Base (TCB), portability across RISC-V platforms, and extensibility to accommodate advanced security features.

The Keystone architecture operates across the multiple privilege levels defined by the RISC-V specification: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). These privilege levels are central to the enforcement of security policies within the system. Keystone's Security Monitor (SM), the heart of its TCB, executes in M-mode and holds exclusive authority over enclave lifecycle management, memory protection configuration, and enforcement of access control policies.

A typical Keystone-enabled platform consists of RISC-V processor cores augmented with secure boot mechanisms, trusted entropy sources, and optional hardware accelerators for cryptographic operations or secure I/O. Keystone does not mandate specific hardware features beyond RISC-V compliance and PMP (Physical Memory Protection), thereby supporting a broad spectrum of implementations from FPGA prototypes to silicon SoCs.

The SM, implemented in C and compiled as a standalone binary, embodies the critical logic responsible for managing the creation, execution, and destruction of enclaves. During enclave creation, the SM validates the enclave's memory layout, initializes its metadata, and configures PMP to ensure that only the SM and the enclave itself can access the designated memory regions. Once enclave execution begins, transitions between host and enclave contexts are mediated by the SM, which dynamically adjusts PMP settings to preserve strict isolation.

Each enclave is composed of two software layers: a user-level application (eapp) and a supervisor-level runtime environment. The eapp contains application-specific logic, while the runtime provides essential operating system abstractions such as exception handling, system call support, and virtual memory management. This two-tiered design allows for flexible enclave development and reduces the complexity of the eapp, which remains entirely untrusted by the host operating system.

The enclave lifecycle encompasses three primary stages: creation, execution, and destruction. During creation, the untrusted host allocates a contiguous region of physical memory—known as Enclave Page Memory (EPM)—to hold the enclave’s code, data, and page tables. This memory is populated and locked by the SM using PMP. In the execution phase, the SM facilitates entry into and exit from the enclave, ensuring that context switches maintain enclave confidentiality. Finally, destruction involves securely erasing enclave memory and releasing resources to prevent post-mortem data leakage.

Keystone also supports optional features such as remote attestation, allowing third-party verifiers to authenticate enclave state before provisioning sensitive data. Additionally, the architecture can be extended to support secure I/O pathways, cryptographic acceleration, and debugging features useful for both development and runtime monitoring.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/keystone_overview.png}
\caption{Overview of the Keystone architecture illustrating components such as the Security Monitor, enclave runtime, and the privilege hierarchy.}
\label{fig:keystone_overview}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=0.9\linewidth]{figures/enclave_lifecycle.png}
\caption{Stages of a Keystone enclave lifecycle: creation, execution, and destruction.}
\label{fig:enclave_lifecycle}
\end{figure}

Overall, Keystone provides a clean, modular, and research-oriented TEE framework that leverages the openness of RISC-V to deliver both strong security guarantees and development flexibility. Its layered architecture, minimal TCB, and standards-compliant interfaces make it a compelling platform for building and evaluating secure applications on next-generation processors.

@@

Keystone distinguishes itself from existing commercial Trusted Execution Environment (TEE) implementations, such as Intel Software Guard Extensions (SGX) and ARM TrustZone, through its open-source nature, customizable architecture, and alignment with the RISC-V ecosystem. While SGX and TrustZone have been widely deployed in commodity hardware and offer robust isolation features, they are constrained by their proprietary designs, large and opaque TCBs, and limited extensibility, which make them less suitable for academic research and fine-grained system-level customization.

ARM TrustZone utilizes a dual-world execution model, dividing the processor into a Secure World and a Normal World. The system transitions between these worlds using secure monitor calls (SMCs). Each world operates with its own set of execution contexts and memory spaces. In the Secure World, a trusted operating system—such as OP-TEE—executes sensitive applications that conform to the GlobalPlatform TEE Internal API. However, TrustZone’s architecture supports only a single secure world instance, lacks support for dynamic enclave creation, and requires significant overhead for world switches, especially in the handling of interrupts. Moreover, the TCB in TrustZone is relatively large, encompassing the entire Secure World kernel and trusted services, which complicates formal verification and increases the risk of vulnerabilities.

Intel SGX, in contrast, supports dynamic creation of enclaves at runtime within the address space of user-space processes. These enclaves are protected by a memory encryption engine and execute in ring 3, isolated from both the operating system and hypervisor. Communication between host and enclave is facilitated by ECALL and OCALL mechanisms, whose interfaces are defined through the edger8r tool. While SGX offers strong isolation and cryptographic sealing of enclave contents, it imposes rigid constraints on enclave memory size and structure. Notably, SGX enclaves do not support supervisor-level execution, limiting the complexity of software that can be securely encapsulated within the enclave. Additionally, SGX is tightly bound to Intel hardware, restricting portability and experimentation.

Keystone integrates the strengths of both TrustZone and SGX while addressing their limitations through an open and modular framework. Its architecture supports dynamic enclave creation, hierarchical privilege separation (via user and supervisor modes), and execution control via a minimal M-mode Security Monitor. Enclave memory is isolated using RISC-V’s Physical Memory Protection (PMP) features, which are dynamically configured by the SM to enforce strict access control. This enables strong spatial and temporal memory isolation without relying on hardware-bound limitations or encryption mechanisms.

Keystone also supports a lightweight communication interface, dubbed keyedger, which serves a similar role to edger8r in SGX, but is designed to be more flexible and extensible. While it currently offers partial support for the GlobalPlatform TEE Internal API, Keystone’s modularity allows it to be extended to support custom APIs and protocols suited to specific use cases or research goals.

Through its combination of openness, modularity, and hardware-agnostic design, Keystone offers a uniquely research-friendly and extensible platform for investigating and deploying secure enclaves. It enables hardware/software co-design, supports verifiable security primitives, and aligns with emerging open hardware ecosystems, positioning it as a forward-looking alternative to incumbent TEE technologies.

### 2025-05-29 chapters/methodology.tex
updated relevant chapters & sections

\section{Experimental setup}

\section{Benchmarking tools and metrics}
%\section{Performance metrics}

@@

>>>>

\section{Experimental setup}
%The experimental setup utilizes the Keystone-enhanced QEMU emulator running on an Ubuntu host operating system. This configuration allows for flexible development and testing of Keystone enclaves in a controlled environment without requiring physical RISC-V hardware. The Ubuntu host provides a stable platform for building and deploying enclave binaries, and it facilitates the integration of the Keystone kernel driver and SDK.

The experimental setup employed in this study was carefully designed to facilitate reliable, repeatable testing of the Keystone Trusted Execution Environment (TEE) without the need for physical RISC-V hardware. To this end, a virtualized testbed was constructed using a RISC-V specific fork of the QEMU emulator, hosted on an Ubuntu 22.04 LTS system. This emulated environment provided a high degree of control over the system configuration, allowing for flexible testing of Keystone enclaves in isolation from physical hardware constraints, while still accurately modeling the behavior of a real RISC-V system.

QEMU, an open-source hardware emulator, was configured to emulate the RV64GC (64-bit general-purpose) RISC-V architecture, which is fully compatible with Keystone’s software stack. The choice of RV64 as the target architecture reflects Keystone’s intended deployment scenario on modern RISC-V platforms and allows full access to the enclave memory protection and privilege separation mechanisms supported by the Security Monitor.

The host operating system—Ubuntu 22.04—was chosen for its long-term support, stable kernel interfaces, and compatibility with the Keystone development toolchain. It served as the platform for building the QEMU binary, compiling the Keystone components, and orchestrating the virtual machine environment in which all benchmarks were executed.

The build process for Keystone relied on a combination of Make and Buildroot. Buildroot, an established tool for generating embedded Linux systems, was used to construct the root filesystem, cross-compile essential system libraries, and configure the Linux kernel used in the virtual guest. The Make-based build system provided streamlined control over component compilation and deployment, enabling reproducible builds of the Keystone runtime, Security Monitor, and associated kernel modules. During the build phase, key configuration parameters were selected, including the choice of QEMU as the target platform, the RV64 architecture, and the specific Buildroot configuration required to support the Keystone kernel driver and enclave runtime environment.

After successful compilation, the Keystone system was launched within QEMU using the command:

\begin{verbatim}
make run
\end{verbatim}

This command initiated the full virtual system, including the Keystone Security Monitor—responsible for managing enclave creation, destruction, and memory isolation—as well as a minimal Linux operating system running in supervisor mode within the QEMU guest. The Linux environment within QEMU was configured to support manual loading of the Keystone kernel driver, which was necessary to enable enclave functionality. Once the system was booted, the kernel driver was inserted using the following command:

\begin{verbatim}
modprobe keystone-driver
\end{verbatim}

This step completed the initialization of the enclave infrastructure, allowing benchmark binaries to be securely loaded into enclaves and executed with isolation enforced by the Security Monitor.

For debugging purposes, additional support was enabled through QEMU’s built-in GDB server. This allowed low-level inspection of the system state, including memory regions, control registers, and exception handlers. When debugging was required, Keystone was recompiled with debugging features enabled by setting the appropriate build flag, and execution was resumed using GDB connectivity:

\begin{verbatim}
KEYSTONE_DEBUG=y make run
make debug-connect
\end{verbatim}

Additional details on the build process, dependencies, and configuration options are available in the \textbf{Appendix}.

\section{Benchmarking tools and metrics}

To evaluate the performance impact of the Keystone Trusted Execution Environment (TEE) in a meaningful and comprehensive way, this study uses two well-established synthetic benchmarks: \textit{Dhrystone} and \textit{CoreMark}. Both are widely used in the embedded systems community and are well-supported in the RISC-V ecosystem, making them a natural fit for assessing how Keystone behaves in secure versus non-secure execution modes. These benchmarks were chosen because they capture a wide range of processor behaviors, particularly those relevant to secure enclave execution—such as integer arithmetic, control flow, and memory access patterns.

The Dhrystone benchmark has been around for decades and was originally designed to measure general-purpose processor performance, especially in environments where floating-point operations and file I/O are not a major concern. Its focus is on integer calculations and logical control structures—things like loops, function calls, and basic data handling—which are typical in many embedded and real-time applications. Because of its simplicity and minimal memory usage, Dhrystone provides a good baseline measure of a processor’s raw computing capability. For this study, the RISC-V version of Dhrystone was obtained from the official \texttt{riscv-tests} repository to ensure that it ran correctly on the target hardware. The benchmark was executed twice under identical conditions—once in a standard, non-secure mode, and once within a Keystone enclave. This setup made it possible to measure the specific performance overhead introduced by Keystone’s security features, such as memory isolation and the cost of switching between secure and non-secure execution contexts.

Dhrystone results are typically expressed in terms of "Dhrystones per second," which simply counts how many iterations of the benchmark are completed in one second. To make the numbers more meaningful across different architectures, these results are often normalized to Dhrystone MIPS (DMIPS), which estimates how many millions of instructions per second the processor is effectively delivering. This helps smooth out differences between instruction sets and gives a more comparable figure. In the context of this study, comparing DMIPS between enclave and non-enclave runs highlights how much raw performance is lost (or retained) when using the TEE.

To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark was also used. Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications. It includes tasks like linked list manipulation, matrix operations, and finite state machine processing, all of which are common in software running on embedded systems. While CoreMark is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.

The version of CoreMark used in this study came from the SiFive repository, which provides a version optimized for RISC-V systems and ready to work with Keystone. Like Dhrystone, CoreMark was run in both native and enclave modes to compare performance under each. Its results are reported in terms of "Iterations Per Second" (IPS), which tells us how many complete runs of the benchmark are executed every second. Since CoreMark includes both arithmetic and memory operations, IPS gives a more holistic view of how the system is performing—not just the processor, but also the memory subsystem.

Beyond the main performance numbers (DMIPS for Dhrystone and IPS for CoreMark), this study also looked at a few supporting metrics to better understand system behavior. The first is overall execution time, which refers to how long it takes (in real-world, wall-clock time) to run each benchmark. Comparing execution time in enclave vs. non-enclave modes helps quantify latency introduced by secure execution. This kind of delay might come from things like extra memory checks, context switching, or other internal mechanisms used to isolate the enclave from the rest of the system.

Another important metric is variability—specifically, the standard deviation of benchmark results across multiple runs. Each benchmark was executed several times under identical conditions, and the standard deviation was calculated to determine how consistent the performance was. Low standard deviation means the system performs reliably from run to run, while high deviation might indicate that performance is sensitive to system noise, scheduling delays, or other environmental factors.

Both Dhrystone and CoreMark are CPU-intensive, which is exactly what makes them useful in this context. Dhrystone focuses narrowly on integer operations and control flow, giving a clean look at raw processor performance. CoreMark, on the other hand, introduces a bit more complexity and memory interaction, which makes it better suited for revealing the broader impact of Keystone’s security mechanisms. Neither benchmark uses file I/O or persistent storage, which means that this study does not explore storage-related features of the TEE, such as secure disk access or file encryption. That’s outside the scope here and would require a different kind of benchmarking approach.

By using both benchmarks together—one simple and focused, the other broader and more realistic—this study creates a well-rounded picture of how Keystone affects performance. The addition of execution time and variability analysis provides further context, helping to separate consistent performance overhead from sporadic slowdowns or system noise. Overall, this methodology provides a clear, detailed view of the trade-offs involved when securing execution through a TEE on RISC-V platforms.

@@

\begin{comment}
In order to conduct a comprehensive and methodologically sound performance evaluation of the Keystone Trusted Execution Environment (TEE), a structured parameter variation strategy was employed. The goal of this strategy was to explore how different system and workload configurations affect performance, both inside and outside of the enclave context. By systematically varying a set of key parameters, it becomes possible to isolate the sources of performance overhead and assess the sensitivity of the system to specific operational conditions.

A fundamental aspect of the evaluation involved comparing benchmark execution in two distinct contexts: native execution versus enclave execution. In the native configuration, benchmarks were executed directly within the emulated Linux environment running on QEMU, without invoking Keystone’s enclave infrastructure. This configuration provided a performance baseline that reflects the raw computational capability of the underlying system without the influence of TEE-specific isolation mechanisms. In contrast, enclave execution involved running the same benchmark binaries within Keystone-enforced secure containers. This allowed for a direct measurement of the overhead introduced by features such as memory isolation, limited system call interfaces, and the additional control flow transitions required by the Security Monitor.

To ensure statistical significance and reproducibility, each benchmark was executed multiple times under identical conditions. Specifically, each configuration (e.g., native or enclave) was run ten times, and the results were aggregated to compute average execution time, throughput (expressed as DMIPS for Dhrystone or IPS for CoreMark), as well as the standard deviation across runs. This repetition helped mitigate the impact of transient system noise or timing anomalies that can occur in virtualized environments, especially under QEMU emulation. Additionally, the inclusion of standard deviation in the reported results provided insights into the consistency and stability of the enclave performance.

Another important axis of variation was workload size. For the CoreMark benchmark, which allows for configurable execution parameters, experiments were conducted using different iteration counts and internal data structure sizes. This enabled the evaluation of how enclave performance scales with increasing computational intensity and memory utilization. Because enclave memory is subject to isolation and access restrictions, increasing the memory footprint can expose performance bottlenecks that would not be apparent under smaller workloads. While Dhrystone does not support direct workload scaling through runtime arguments, multiple versions of the benchmark were compiled with different loop unroll factors to simulate variation in workload complexity. Although less precise than CoreMark’s controls, this approach still provided a meaningful indication of how performance trends shift with increasing instruction throughput requirements.

The influence of system-level configuration options was also examined. Keystone offers various compile-time and runtime flags that affect behavior and performance, most notably the debug mode. When enabled, debug mode activates additional runtime checks, logging mechanisms, and diagnostic outputs intended to aid developers during enclave development and testing. These features, while useful during implementation, introduce measurable overhead. By executing benchmarks with and without debug mode enabled, the experiments were able to quantify the cost of enhanced observability and determine whether such features significantly degrade performance.

Memory allocation parameters were also varied to understand their role in shaping enclave performance. In particular, the heap and stack sizes assigned to benchmarks were adjusted to evaluate the sensitivity of the system to memory layout and availability. This was particularly relevant for CoreMark, which dynamically allocates and manipulates internal data structures. Larger memory allocations could potentially lead to increased cache pressure or page faults within the enclave, highlighting the importance of efficient memory management in TEE contexts.

Measurement accuracy was further reinforced by cross-validating timing results. Execution time was captured using both the benchmarks’ internal timing functions and QEMU’s emulated timer infrastructure. This dual-timing approach ensured that timing results were not distorted by emulator artifacts or low-resolution timers. In cases where discrepancies were observed, additional investigation was conducted to determine the root cause, and only consistent, corroborated results were included in the final analysis.

Finally, to explore the potential impact of system-level contention, a set of exploratory experiments introduced artificial background load during native execution. Although Keystone enclaves currently operate in a single-threaded, non-preemptive mode, testing under load offered preliminary insights into how enclave performance might be influenced in future multi-threaded or multi-tenant scenarios, especially in shared virtualized environments. While these tests were not a primary focus of the evaluation, they provided valuable context for interpreting variability in the results.
\end{comment}

### 2025-05-29 main.tex
updated relevant chapters & sections

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{computer modern oldschool design}
\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here

@@

\chapter{Introduction}
\input{chapters/01_introduction.tex}

\chapter{Background and System Architecture}
\input{chapters/02_background.tex}

\chapter{Methodology}
\input{chapters/03_methodology.tex}

\chapter{Results and discussion}
\input{chapters/04_results.tex}

\chapter{System Configuration Recommendations}
\input{chapters/05_recommendations.tex}

\chapter{Conclusion and Future Work}
\input{chapters/06_conclusion.tex}

\begin{bibtex-entries}
@TechReport{Kernighan1974,
author = {Brian Kernighan},
title = {Programming in C – A Tutorial},
institution = {Bell Laboratories},
year = {1974}
}
\end{bibtex-entries}
\end{document}

>>>>

\documentclass[english, version-2020-11]{uzl-thesis}
\UzLStyle{computer modern oldschool design}
\usepackage{comment}
\usepackage{float}

\input{preamble.tex} % Put your \UzLStyle and \UzLThesisSetup here

@@

\chapter{Introduction}

Modern computing increasingly relies on platforms that operate in potentially untrusted environments. These include public cloud infrastructures, multi-tenant servers, and distributed edge devices, where sensitive data and critical computations may be exposed to compromised software stacks or malicious actors. In such contexts, ensuring the confidentiality and integrity of both data and execution is a fundamental requirement. Conventional security mechanisms, such as access control and encryption, are often insufficient in scenarios where the underlying operating system, hypervisor, or even firmware may be compromised.

To address these challenges, Trusted Execution Environments (TEEs) have emerged as a compelling solution. TEEs provide hardware-assisted isolation mechanisms that enable the execution of code in a protected context, referred to as an enclave. This enclave is designed to be resilient to attacks originating from untrusted system components, including the operating system and other user-space applications. By establishing a secure boundary around sensitive code and data, TEEs allow developers to implement trustworthy services even on compromised hosts.

Several commercial implementations of TEEs exist today, with Intel Software Guard Extensions (SGX) and ARM TrustZone being the most prominent examples. These technologies have demonstrated the practical viability of hardware-enforced isolation, and they have been integrated into a variety of real-world applications, from digital rights management to secure machine learning inference. However, despite their utility, these solutions are inherently proprietary and closed-source. This lack of transparency imposes significant constraints on researchers, developers, and system architects who wish to investigate alternative TEE designs, customize security mechanisms, or perform formal verification. Moreover, the architectural rigidity of these platforms limits their adaptability to emerging use cases, especially in academic or experimental contexts.

To overcome these limitations, the Keystone framework introduces an open-source, modular TEE architecture built atop the RISC-V instruction set architecture (ISA). RISC-V itself is an open, extensible ISA that has gained significant traction in both academic and industrial settings. Keystone leverages RISC-V's flexibility to enable fine-grained control over enclave behavior and to support a minimal, verifiable trusted computing base (TCB). The framework allows researchers and developers to explore custom enclave policies, experiment with hardware-software co-design, and perform reproducible evaluations in a controlled setting.

This thesis presents a detailed investigation into the Keystone TEE framework. It explores Keystone's architectural foundations, its use of RISC-V hardware features such as Physical Memory Protection (PMP), and its support for dynamic enclave management. Special attention is paid to the performance implications of enclave isolation, as measured through controlled benchmarking in a virtualized environment. By evaluating Keystone under varying execution conditions, this work seeks to provide a rigorous assessment of its efficiency and suitability for secure computation on emerging RISC-V platforms.
\input{chapters/introduction.tex}

\chapter{Background and System Architecture}
The Keystone framework is built on the RISC-V instruction set architecture, which provides a unique foundation for secure system design due to its open specification and modular extensibility. Unlike traditional ISAs that are controlled by single vendors, RISC-V encourages community-driven enhancements and supports custom extensions, making it particularly attractive for trusted computing research and development. Its clean and well-documented privilege model allows for the implementation of isolation mechanisms that are critical to the construction of TEEs.

RISC-V defines three primary privilege levels: user mode (U-mode), supervisor mode (S-mode), and machine mode (M-mode). These privilege levels are hierarchical, with M-mode having the highest authority over system resources. M-mode has direct access to all hardware features and is responsible for critical functions such as interrupt control, exception handling, and access to physical memory protection mechanisms. Keystone leverages this privilege model by placing its Security Monitor (SM)—the core of the TCB—in M-mode. The SM is responsible for managing enclaves, enforcing memory isolation, and mediating access to sensitive system operations. By situating the TCB in M-mode, Keystone ensures that enclave isolation policies are enforced at the highest level of privilege, thereby minimizing the attack surface.

A central hardware feature used by Keystone is Physical Memory Protection (PMP), introduced in version 1.10 of the RISC-V Privilege Specification. PMP enables M-mode software to define access permissions for specific physical memory regions, controlling which lower-privilege modes (U and S) may read, write, or execute code in those regions. This capability forms the basis of Keystone’s memory isolation model. Each enclave is allocated a private memory region protected by PMP, ensuring that neither the host operating system nor any unauthorized software can access its contents. PMP rules are enforced directly by hardware, providing strong guarantees of isolation that are resistant to software-based attacks.

In addition to memory protection, RISC-V supports a flexible trap and exception handling mechanism. M-mode can intercept and handle all traps, but may delegate certain classes of exceptions—such as system calls and page faults—to S-mode for performance optimization. This delegation allows the enclave runtime to cooperate with the host operating system for memory management tasks, while still maintaining strict control over enclave boundaries. The virtual memory subsystem is managed through the standard RISC-V MMU, which translates virtual addresses to physical addresses using multi-level page tables. Keystone leverages this infrastructure to support enclaves that operate with their own private address spaces, protected by PMP from tampering or inspection.

Importantly, the use of standard programming models and toolchains for M-mode software development makes the Keystone Security Monitor both maintainable and extensible. Unlike microcode or hardwired logic, which are difficult to audit and update, the SM is implemented in C and can be subjected to conventional testing and formal verification techniques. This increases the transparency and trustworthiness of the TCB, which is a critical requirement for secure system design.

Taken together, the combination of RISC-V’s privilege hierarchy, PMP, trap delegation mechanisms, and extensibility provides a solid foundation for implementing robust and configurable TEEs. The Keystone framework capitalizes on these features to deliver a flexible and open platform for secure enclave execution, while remaining faithful to the RISC-V philosophy of openness and modularity.

\input{chapters/background.tex}

\chapter{Methodology}
The primary objective of this thesis is to assess the performance impact of Keystone’s enclave isolation mechanisms on typical embedded and systems-level workloads. Specifically, the study seeks to quantify the computational overhead introduced by executing applications within a Keystone enclave, as compared to their execution in a conventional, non-isolated environment. To achieve this, a series of carefully controlled experiments were designed and executed within a virtualized RISC-V system based on QEMU.

All benchmarks were executed in an emulated environment configured for the RV64GC architecture, which is representative of 64-bit RISC-V systems targeted by Keystone. The use of QEMU for emulation allowed for fine-grained control over the system environment, ensured repeatability, and eliminated the variability introduced by physical hardware, such as thermal throttling or hardware-specific optimizations. Although virtualized, the emulation accurately models instruction execution, privilege transitions, memory access patterns, and I/O behavior, making it suitable for preliminary performance evaluation.

To evaluate the impact of enclave execution, each benchmark was implemented as a standalone enclave application (referred to as an “eapp”), accompanied by a corresponding host application that manages enclave lifecycle events. The host application, running in user space, is responsible for initializing the enclave, loading the benchmark binary, invoking enclave entry points, and handling edge calls used for communication between the enclave and the host. The enclave application, in contrast, is isolated by the Keystone runtime and contains only the benchmark logic and internal data structures. This modular separation ensures that the same benchmark code can be executed both natively (without an enclave) and securely (within an enclave) without structural modification, enabling an accurate comparative analysis.

The evaluation involved executing each benchmark—Dhrystone and CoreMark—under two distinct configurations:

\begin{enumerate}
\item \textit{Native Execution:} The benchmark runs as a conventional user-space process within the QEMU-emulated Linux system, without invoking Keystone enclave services.
\item \textit{Enclave Execution:} The benchmark is loaded into a Keystone enclave, with isolation enforced by the Security Monitor using PMP.
\end{enumerate}

Each benchmark was run multiple times (typically ten iterations per configuration) to account for performance variability and ensure statistical robustness. For each run, metrics such as total execution time, throughput (measured in DMIPS for Dhrystone and iterations per second for CoreMark), and standard deviation were collected. The comparative analysis focused on the relative performance degradation observed in the enclave configuration, thereby providing a direct measurement of the cost of security.

This methodology enables a nuanced understanding of how Keystone’s isolation features influence real-world performance metrics. The controlled nature of the test environment, coupled with repeated measurement and the use of industry-standard benchmarks, ensures that the results are both meaningful and reproducible. By isolating the effect of enclave execution, this study contributes valuable insights into the trade-offs between security and efficiency in open-source TEE architectures.
\input{chapters/methodology.tex}

\chapter{Results and discussion}
\input{chapters/results.tex}

\chapter{System Configuration Recommendations}
\input{chapters/recommendations.tex}

\chapter{Conclusion}
%\input{chapters/conclusion.tex}

\begin{bibtex-entries}
@TechReport{Kernighan1974,
author = {Brian Kernighan},
title = {Programming in C – A Tutorial},
institution = {Bell Laboratories},
year = {1974}
}

@article{suzaki2021ts,
  title={Ts-perf: General performance measurement of trusted execution environment and rich execution environment on intel sgx, arm trustzone, and risc-v keystone},
  author={Suzaki, Kuniyasu and Nakajima, Kenta and Oi, Tsukasa and Tsukamoto, Akira},
  journal={IEEE Access},
  volume={9},
  pages={133520--133530},
  year={2021},
  publisher={IEEE}
}

\end{bibtex-entries}
\end{document}
